---
title: "What Drives Rent?"
subtitle: "A Deep Dive Into People, Place, and Price"
output:
  pdf_document: 
    latex_engine: xelatex
    includes:
      in_header: header.tex
---


```{r library}
#| echo: false
#| warning: false
#| include: false


library(tidyverse)
library(ggplot2)
library(dplyr)
library(openxlsx)
library(stargazer)
library(scales)
library(sandwich)

```


## Introduction

This analysis investigates how demographic and socioeconomic factors influence one-bedroom rental prices across five U.S. regions: Northeast, Northwest, Southeast, Southwest, and Midwest. Using ZIP Code–level data merged from ACS, HUD FMR, and HUD-USPS crosswalk files, region-specific linear regression models with robust standard errors were applied to identify significant predictors of rent variation. The goal is to quantify localized effects of variables such as race, income, education, and marital status on rental housing costs.


## Description of the Data Source


Data sources from three publicly available sources were used in this analysis to investigate how demographic and socioeconomic factors relate to rental prices.

1. American Community Survey (ACS) 
The U.S. Census Bureau conducts the ACS annually to provide detailed demographic, social, economic, and housing data. Approximately 3.5 million addresses are randomly sampled each year, with participation required by law, ensuring a high response rate and robust data quality. The unit of observation in this dataset is the ZIP Code. We used the 2023 ACS S0601 dataset, which includes variables such as median income, educational attainment, racial composition, and housing characteristics, which are core indicators for examining disparities across regions.

2. Fair Market Rent (FMR)
Rent data is derived from the 2023 Fair Market Rent (FMR) published annually by the U.S. Department of Housing and Urban Development (HUD), through its Office of Policy Development and Research, for use in federal housing programs such as Section 8. These rent estimates are calculated using a combination of the Census Bureau’s five-year ACS data (as a base), HUD-defined geographic areas, and adjustments based on the Bureau of Labor Statistics’ Consumer Price Index (CPI). The unit of observation here is also at the ZIP-level, aligning with the ACS data for comparison. FMR serves as a standardized estimate of local rental costs, allowing us to assess rent relative to regional socioeconomic conditions.

3. HUD-USPS ZIP-TRACT Crosswalk File
Residential vacancy indicators are derived from ZIP-Track Crosswalk files produced by HUD in collaboration with the U.S. Postal Service. The 2023 Q2 file used filters out ZIP Codes with low residential ratios or primarily commercial/industrial land use, allowing the analysis to focus on deriving meaningful insights from relevant areas.

To ensure balanced sampling across geographic regions, the dataset was stratified into five regions (Midwest, Northeast, Southeast, Southwest, and Northwest) and 1,000 ZIP Codes were randomly selected per region, resulting in a final dataset of 5,000 rows and 37 columns. This consolidated dataframe served as the single source of truth for all subsequent modeling and visualization steps.


```{r data preparation}
#| echo: false
#| warning: false
#| include: false

## Data Processing

data = read.csv("data/dataset.csv")

data = data.frame(data) %>%
  mutate(across(c(
    Median.Age,
    Percent.less.than.high.school,
    Percent.some.college.or.associates,
    Percent.college.graduate,
    Percent.graduate.degree,
    Median.income.in.past.year,
    Percent..100.poverty.status,
    Percent..100.TO.149.poverty.status,
    Percent..150.poverty.status,
    Percent.high.school.graduate
  ), as.numeric))

```


## Linear Regression Models

To examine how demographic and socioeconomic factors relate to rent prices across the United States, five separate linear regressions were conducted—one for each region: Northeast (NE), Northwest (NW), Southeast (SE), Southwest (SW), and Midwest (MW). Each model used the SAFMR 1BR variable (Small Area Fair Market Rent for one-bedroom units) as the dependent variable, with demographic indicators such as race, marital status, educational attainment, income, and language use as predictors.

The tables display the full regression results with robust standard errors (HC1) to address potential heteroskedasticity. In the following sections, most notable results from each region are summarized, focusing on statistically significant coefficients (p < 0.05), the direction and magnitude of associations, and any unique patterns that emerge across geographies. 



### Regression Results for the Northeastern Region

In the Northeast, higher percentages of Black residents and Hispanic/Latino populations are associated with significantly lower rent prices, with coefficients of -80.33 and -67.40 respectively (both p < 0.01). Conversely, a higher percentage of Asian residents is associated with higher rents (+67.64, p < 0.05). Median income also shows a strong positive effect on rent (+0.13 per $1,000, p < 0.01). Educational attainment, marital status, and other demographic variables showed no significant associations in this region.


```{r lm by region NE, results='asis'}
#| echo: false
#| warning: false


model_NE = lm(SAFMR.1BR ~ Median.Age+Percent.Males+Percent.Females+
                Percent.Race..White+Percent.Race..Black+Percent.Race..Native.or.Alaskan+Percent.Race..Asian+Percent.Race..Native.Hawaiian.or.Pacific.Islander+Percent.Race..Hispanic.or.Latino+Percent.speak.another.language.at.home+
                Percent.never.married+Percent.married+Percent.divorced+Percent.widowed+Percent.Race..Two.or.more.races+
                Percent.less.than.high.school+Percent.high.school.graduate+Percent.some.college.or.associates+Percent.college.graduate+Percent.graduate.degree+
                Median.income.in.past.year+Percent.born.in.state.of.residence+Percent..100.poverty.status,
    data = data[data$Region == 'NE',])
df_NE = data.frame(summary(model_NE)$coefficients)
df_NE = df_NE %>%
  mutate(variables = rownames(df_NE), Region = 'NE')
rownames(df_NE) = 1:nrow(df_NE)


robust_vcov <- vcovHC(model_NE, type = "HC1")

stargazer(model_NE, type = "latex", title = "Regression Results for NE Region", 
          se = list(sqrt(diag(robust_vcov))),
          dep.var.labels = "SAFMR 1BR",
          column.sep.width = "-2pt",  
          font.size = "scriptsize",      
          single.row = TRUE,
          table.layout = "-d#-t-a-s-n",  
          style = "default",
          out.header = FALSE)

```


### Regression Results for the Northwestern Region

In the Northwest, rents rise significantly with Median Age (+11.99, p < 0.01) and Median Income (+0.07 per $1,000, p < 0.01). Black (%) and Hispanic/Latino (%) show significant negative associations with rent (-77.70 and -72.97 respectively, p < 0.01). Interestingly, Percent Married has a large positive effect on rent (+74.51, p < 0.01), suggesting a link between household structure and rental prices in the NW region.


```{r lm by region NW, results='asis'}
#| echo: false
#| warning: false

model_NW = lm(SAFMR.1BR ~ Median.Age+Percent.Males+Percent.Females+
                Percent.Race..White+Percent.Race..Black+Percent.Race..Native.or.Alaskan+Percent.Race..Asian+Percent.Race..Native.Hawaiian.or.Pacific.Islander+Percent.Race..Hispanic.or.Latino+Percent.speak.another.language.at.home+
                Percent.never.married+Percent.married+Percent.divorced+Percent.widowed+Percent.Race..Two.or.more.races+
                Percent.less.than.high.school+Percent.high.school.graduate+Percent.some.college.or.associates+Percent.college.graduate+Percent.graduate.degree+
                Median.income.in.past.year+Percent.born.in.state.of.residence+Percent..100.poverty.status,
              data = data[data$Region == 'NW',])
df_NW = data.frame(summary(model_NW)$coefficients)
df_NW = df_NW %>%
  mutate(variables = rownames(df_NW), Region = 'NW')
rownames(df_NW) = 1:nrow(df_NW)


robust_vcov <- vcovHC(model_NW, type = "HC1")


stargazer(model_NW, type = "latex", title = "Regression Results for NW Region", 
          se = list(sqrt(diag(robust_vcov))),
          dep.var.labels = "SAFMR 1BR",
          column.sep.width = "-2pt",  
          font.size = "scriptsize",      
          single.row = TRUE,
          table.layout = "-d#-t-a-s-n",  
          style = "default",
          float = TRUE,
          out.header = FALSE)

```

### Regression Results for the Southeastern Region

In the Southeast, Percent Asian residents have a positive and significant association with rent prices (+84.75, p < 0.01). In contrast, Percent Hispanic/Latino and Percent with Less than High School education are both linked to lower rents (-59.58 and -61.97 respectively, p < 0.01). Percent Married again shows a strong positive link to higher rents (+71.86, p < 0.01), consistent with patterns observed in the Northwest.


```{r lm by region SE, results='asis'}
#| echo: false
#| warning: false

model_SE = lm(SAFMR.1BR ~ Median.Age+Percent.Males+Percent.Females+
                Percent.Race..White+Percent.Race..Black+Percent.Race..Native.or.Alaskan+Percent.Race..Asian+Percent.Race..Native.Hawaiian.or.Pacific.Islander+Percent.Race..Hispanic.or.Latino+Percent.speak.another.language.at.home+
                Percent.never.married+Percent.married+Percent.divorced+Percent.widowed+Percent.Race..Two.or.more.races+
                Percent.less.than.high.school+Percent.high.school.graduate+Percent.some.college.or.associates+Percent.college.graduate+Percent.graduate.degree+
                Median.income.in.past.year+Percent.born.in.state.of.residence+Percent..100.poverty.status,
              data = data[data$Region == 'SE',])
df_SE = data.frame(summary(model_SE)$coefficients)
df_SE = df_SE %>%
  mutate(variables = rownames(df_SE), Region = 'SE')
rownames(df_SE) = 1:nrow(df_SE)


robust_vcov <- vcovHC(model_SE, type = "HC1")

stargazer(model_SE, type = "latex", title = "Regression Results for SE Region", 
          se = list(sqrt(diag(robust_vcov))),
          dep.var.labels = "SAFMR 1BR",
          column.sep.width = "-2pt",  
          font.size = "scriptsize",      
          single.row = TRUE,
          table.layout = "-d#-t-a-s-n",  
          style = "default",
          out.header = FALSE)

```


### Regression Results for the Southwestern Region

The Southwest model shows strong negative associations between rent and both Percent Hispanic/Latino (-75.64, p < 0.01) and Percent Less Than High School (-73.78, p < 0.01). Median Income remains a significant positive predictor (+0.11 per $1,000, p < 0.01). Like in the NW and SE, Percent Married has a positive and statistically significant effect on rent (+68.36, p < 0.01).


```{r lm by region SW, results='asis'}
#| echo: false
#| warning: false

model_SW = lm(SAFMR.1BR ~ Median.Age+Percent.Males+Percent.Females+
                Percent.Race..White+Percent.Race..Black+Percent.Race..Native.or.Alaskan+Percent.Race..Asian+Percent.Race..Native.Hawaiian.or.Pacific.Islander+Percent.Race..Hispanic.or.Latino+Percent.speak.another.language.at.home+
                Percent.never.married+Percent.married+Percent.divorced+Percent.widowed+Percent.Race..Two.or.more.races+
                Percent.less.than.high.school+Percent.high.school.graduate+Percent.some.college.or.associates+Percent.college.graduate+Percent.graduate.degree+
                Median.income.in.past.year+Percent.born.in.state.of.residence+Percent..100.poverty.status,
              data = data[data$Region == 'SW',])
df_SW = data.frame(summary(model_SW)$coefficients)
df_SW = df_SW %>%
  mutate(variables = rownames(df_SW), Region = 'SW')
rownames(df_SW) = 1:nrow(df_SW)


robust_vcov <- vcovHC(model_SW, type = "HC1")

stargazer(model_SW, type = "latex", title = "Regression Results for SW Region", 
          se = list(sqrt(diag(robust_vcov))),
          dep.var.labels = "SAFMR 1BR",
          column.sep.width = "-2pt",  
          font.size = "scriptsize",      
          single.row = TRUE,
          table.layout = "-d#-t-a-s-n",  
          style = "default",
          out.header = FALSE)

```


### Regression Results for the Midwest Region

In the Midwest, Black (%) is associated with significantly lower rents (-62.41, p < 0.01), as is Percent Less Than High School (-65.66, p < 0.01). The positive association between Median Income and rent persists here (+0.12 per $1,000, p < 0.01). Unique to this region, Percent Native Hawaiian or Pacific Islander has a significant positive coefficient (+318.45, p < 0.01), though the sample size for this group may be small.


```{r lm by region MW, results='asis'}
#| echo: false
#| warning: false

model_MW = lm(SAFMR.1BR ~ Median.Age+Percent.Males+Percent.Females+
                Percent.Race..White+Percent.Race..Black+Percent.Race..Native.or.Alaskan+Percent.Race..Asian+Percent.Race..Native.Hawaiian.or.Pacific.Islander+Percent.Race..Hispanic.or.Latino+Percent.speak.another.language.at.home+
                Percent.never.married+Percent.married+Percent.divorced+Percent.widowed+Percent.Race..Two.or.more.races+
                Percent.less.than.high.school+Percent.high.school.graduate+Percent.some.college.or.associates+Percent.college.graduate+Percent.graduate.degree+
                Median.income.in.past.year+Percent.born.in.state.of.residence+Percent..100.poverty.status,
              data = data[data$Region == 'MW',])
df_MW = data.frame(summary(model_MW)$coefficients)
df_MW = df_MW %>%
  mutate(variables = rownames(df_MW), Region = 'MW')
rownames(df_MW) = 1:nrow(df_MW)

robust_vcov <- vcovHC(model_SW, type = "HC1")

stargazer(model_MW, type = "latex", title = "Regression Results for MW Region", 
          se = list(sqrt(diag(robust_vcov))),
          dep.var.labels = "SAFMR 1BR",
          column.sep.width = "-2pt",  
          font.size = "scriptsize",      
          single.row = TRUE,
          table.layout = "-d#-t-a-s-n",  
          style = "default",
          out.header = FALSE)

```

```{r join results}
#| echo: false


df = rbind(df_NE, df_SE, df_SW, df_NW, df_MW) %>%
  mutate(p_value = Pr...t.., std_error = Std..Error) %>% # Rename p_value and std_error
  select(variables, Region, Estimate, std_error, t.value, p_value) %>%
  mutate(Estimate = case_when(variables == "Median.income.in.past.year" ~ Estimate*1000, TRUE ~ Estimate), 
         Sig = case_when(p_value < 0.001 ~ '***',  # Create significance indicator based on p_value
                         between(p_value, 0.001, 0.01) ~ '**',
                         between(p_value, 0.01, 0.05) ~ '*',
                         between(p_value, 0.05, 0.1) ~ '.',
                         TRUE ~ ''))

```

## Graph

### Demographic and Geographic Impact on Rental Prices

The graph illustrates the estimated change in rental prices associated with one-unit increases in various demographic and socioeconomic indicators, based on region-specific linear regression models. Each bar represents a coefficient from the model, color-coded by statistical significance (from p < 0.1 being lightest to p < 0.001 being darkest), and measured in dollars per unit increase in the variable.

Across all five regions, median income is the most consistently impactful variable, showing a strong and statistically significant positive association with rent prices. A \$1,000 increase in median income within a ZIP Code is associated with a \$60 to \$120 increase in the SAFMR 1-bedroom rent estimate. 

Overall, the graph highlights that while income is a strong universal predictor of rent, the influence of race, education, and marital structure varies meaningfully by region. The presence of consistent negative rent associations for Black and Hispanic/Latino populations underscores persistent structural disparities, even after controlling for income and other demographic variables. Meanwhile, positive associations with educational attainment and marriage suggest that perceived indicators of socioeconomic status are embedded in how rent is priced at the neighborhood level.

```{r plot, fig.height=7, fig.width=10}
#| echo: false


plot1 = df %>%
  select(Region, Estimate, Sig, variables) %>%
  mutate(var= case_when(variables == "Median.Age" ~ "Median Age",
                     variables == "Percent.Males" ~ "Male (%)",
                     variables == "Percent.Race..Black" ~ "Black (%)",
                     variables == "Percent.Race..Native.or.Alaskan" ~ "Native/Alaskan (%)",
                     variables == "Percent.Race..Asian" ~ "Asian (%)",
                     variables == "Percent.Race..Native.Hawaiian.or.Pacific.Islander" ~ "Hawaiian/Pacific Islander (%)",
                     variables == "Percent.Race..Hispanic.or.Latino" ~ "Hispanic/Latino (%)",
                     variables == "Median.income.in.past.year" ~ "Median Income ($000)",
                     variables == "Percent.Race..Two.or.more.races" ~ "2+ Race (%)",
                     variables == "Percent.less.than.high.school" ~ "Less Than High School (%)",
                     variables ==  "Percent.married" ~ "Married (%)",
                     variables ==  "Percent.born.in.state.of.residence" ~ "Born in Residing State (%)",
                     variables ==  "Percent..100.poverty.status" ~ "Below 100% Poverty Level (%)",
                     TRUE ~ ""))

plot1 %>% 
  filter(var %in% c("Median Age",
                     "Median Income ($000)",
                     "Male (%)",
                     "Below 100% Poverty Level (%)",
                     "Black (%)",
                     "2+ Race (%)",
                     "Asian (%)",
                     "Hispanic/Latino (%)",
                    "Less Than High School (%)",
                    "Married (%)",
                    "Born in Residing State (%)"
                    )) %>%

  ggplot(aes(x=Region, y=Estimate, fill=factor(Sig, levels=c('', '.', '*', '**', '***'))))+
  scale_fill_manual(values=c("#e0d7d7", "#ebc3c3", "#d47979","#d63a3a", "#990000")) +
  facet_wrap(vars(var), ncol=3) +
  geom_col(position="dodge")+
  coord_flip() + 
  geom_hline(yintercept=0) +
  labs(title="Demographic and Geographic Impact on Rental Prices", x="", y="Change in Rental Price ($) Per Unit Increase", fill="Stat. Sig.") +
  scale_y_continuous(trans='pseudo_log', breaks = c(-100, -10,0, 10,100, 1000))


```



## Citation


U.S. Census Bureau, “S0601 Selected Characteristics of the Total and Native Populations in the United States”, <https://data.census.gov/table?q=s0601>, accessed on March 25, 2025.

HUD User’s Office of Policy Development and Research, “Fair Market Rents”, <www.huduser.gov/portal/datasets/fmr.html>, accessed on March 25, 2025.

HUD User’s Office of Policy Development and Research, “HUD-USPS ZIP Crosswalk Files” <www.huduser.gov/apps/public/uspscrosswalk/home>, accessed on March 25, 2025.